---
title: "PupilBio Challenge"
output:
  pdf_document: default
  html_notebook: default
---

```{r Loading libraries,message=FALSE}
library(readr)
library(ggplot2)
library(caTools)
library(xgboost)
library(caret)
library(tibble)
```


Presets and reading input
```{r read input files,message=FALSE}
pmp <- read_csv("../../Downloads/PupilBioTest_PMP_revA.csv", col_names = T)
cfdna <- pmp[pmp$Tissue == "cfDNA",]
islet <- pmp[pmp$Tissue == "Islet",]
```

Function to calculate coverage of both tissues.

```{r defined functions}
calculate_coverage <- function(df){
  to_return <- data.frame()
  for (coord in unique(df$CpG_Coordinates)) {
    each_cpg <- df[df$CpG_Coordinates == coord,]
    per_sample_rep <- rowSums(each_cpg[,c(3:10)])
    #Coverage
    coverage <- sum(per_sample_rep)
    #Coefficient of Variance
    cv <- sd(per_sample_rep)/mean(per_sample_rep)*100
    #Median
    median <- median(per_sample_rep)
    to_return <- rbind(to_return,data.frame(CpG_Coordinates = coord,
                                            Coverage = coverage, 
                                            Tissue = unique(each_cpg$Tissue),
                                            Median = median, CV=cv))
  }
  return(to_return)
}
```


Calculaion of said coverage
```{r Get coverage,message=FALSE}
#This is the actual code but my R keeps crashing so I'm reading the files directly
#islet_coverage <- calculate_coverage(df = islet)
#summary(islet_coverage)
#write_csv(islet_coverage,"projects/side_project/islet_coverage.csv", col_names = T)
#cfdna_coverage <- calculate_coverage(df = cfdna)
#summary(cfdna_coverage)
#write_csv(cfdna_coverage,"projects/side_project/cfDNA_coverage.csv", col_names = T)

islet_coverage <- read_csv("../../islet_coverage.csv", col_names = T)
cfdna_coverage <- read_csv("../../cfDNA_coverage.csv", col_names = T)
summary(islet_coverage)
summary(cfdna_coverage)
```

Based on the box plot, it is quite apparent that the cfDNA has much higher coverage than Islet cell.
```{r Plotting}
plot_df <- rbind(islet_coverage,cfdna_coverage)
ggplot(plot_df, aes(x=Tissue, y=Coverage, fill=Tissue)) + geom_boxplot()
ggplot(plot_df, aes(x=Coverage, fill = Tissue)) + geom_histogram()
```

Removing PMP with coverage above 400000 to better visualize the data (by removing the extremes)

```{r Plotting 2}
cfdna_coverage <- cfdna_coverage[cfdna_coverage$Coverage <= 400000,]
plot_df <- rbind(islet_coverage,cfdna_coverage)
ggplot(plot_df, aes(x=Tissue, y=Coverage, fill=Tissue)) + geom_boxplot()

#Then further decreasing to 200000 to get a granular read.
cfdna_coverage <- cfdna_coverage[cfdna_coverage$Coverage <= 200000,]
plot_df <- rbind(islet_coverage,cfdna_coverage)
ggplot(plot_df, aes(x=Tissue, y=Coverage, fill=Tissue)) + geom_boxplot()
```

A gradient boosting model was used to predict the classification of this dataset using xgboost. Because the dataset is heavily skewed towards cfDNA(0) with 75% of data to Islet cells' 25%, the imbalance is corrected by heavily penalizing errors in the minority class (Islet cell) xgboost (gradient boosting) and thus setting scale_pos_weight to 3.

If I had the computational bandwidth, I would have done a k-fold cross validation.

```{r Classification}
#Splitting dataset into test:training in 25:75 ratio at random (and thus setting seed)
pmp$Tissue <- ifelse(pmp$Tissue == "Islet",1,0)
set.seed(234)
split <- sample.split(pmp$Tissue, SplitRatio = 0.75)
training_set <- subset(pmp[,c(3:10,13)], split == TRUE)
training_set_pmp <- subset(pmp[,2], split == TRUE)
test_set <- subset(pmp[,c(3:10,13)], split == FALSE) 
test_set_pmp <- subset(pmp[,2], split == FALSE)

# Prepare data matrices
train_matrix <- as.matrix(training_set[,-9])
test_matrix <- as.matrix(test_set[,-9])
train_label <- training_set$Tissue
test_label <- test_set$Tissue


#Gradient Boosting model, prediction and associated statistics
xgb_model <- xgboost(verbose = F,data = train_matrix, 
                     label = train_label, objective = "binary:logistic", 
                     nrounds = 100, scale_pos_weight = 3)
xgb_predictions <- predict(xgb_model, test_matrix)
xgb_rmse <- sqrt(mean((xgb_predictions - test_label)^2))
print(paste("RMSE (XGBoost):", round(xgb_rmse, 2)))
importance <- xgb.importance(feature_names = colnames(train_matrix), model = xgb_model)
xgb.plot.importance(importance_matrix = importance)
saveRDS(xgb_model,"xgb_n100_scaleposweight3.rds")
```



```{r Predictions}
test_set$probability <- xgb_predictions
test_set$prediction <- ifelse(test_set$probability > 0.5,1,0)
test_set$Tissue <- factor(test_set$Tissue)
test_set$prediction <- factor(test_set$prediction)
test_set_pmp <- cbind(test_set_pmp,test_set)
test_set_pmp$coverage <- rowSums(test_set_pmp[,c(2:9)])
write_csv(test_set_pmp,"entire_test_set.csv", col_names = T)
```


```{r Confusion Matrix}
confusion_mat <- confusionMatrix(test_set$Tissue,test_set$prediction)
print(confusion_mat)
tocsv <- data.frame(cbind(t(confusion_mat$overall),t(confusion_mat$byClass)))
temp <- data.frame(t(tocsv))
temp <- rownames_to_column(temp,"Statistic")
names(temp)[2] <- "Value"
write_csv(temp,file="Statistics_scaleweightpos3_n100.csv",col_names = T)
write_csv(as.data.frame(confusion_mat$table),file="confusionMatrix_scaleweightpos3_n100.csv",col_names = T)
rm(temp,tocsv)
```
Conclusion: As intended, the FP was minimized while allowing some FN.


```{r Mean Variant Fraction Calculation}
#This is the actual code, but because of memory issues I'm reading directly from the file
#mean_variant_fraction <- data.frame()
#for (coord in unique(pmp$CpG_Coordinates)) {
#  each_pmp <- pmp[pmp$CpG_Coordinates == coord,]
#  total_reads_per_variant <- colSums(each_pmp[,c(3:10)])
#  total_reads <- sum(total_reads_per_variant)
#  mvf <- total_reads_per_variant/total_reads
#  each_pmp <- c(CpG_Coordinates = coord,mvf)
#  mean_variant_fraction <- rbind(mean_variant_fraction,each_pmp)
#}
#colnames(mean_variant_fraction) <- c("CpG_Coordinates","X000","X001","X010","X011","X100","X101","X110","X111")
#mean_variant_fraction[,c(2:9)] <- apply(mean_variant_fraction[,c(2:9)],2,as.numeric)
#write_csv(mean_variant_fraction,"mean_variant_fraction.csv", col_names = T)
mean_variant_fraction <- read_csv("mean_variant_fraction.csv", col_names = T)
summary(mean_variant_fraction)
```


3a) Specificity is a measure of true negative (prediction accuracy of Tissue #2). In this case, since Tissue 2 has significantly low coverage, our ability to get a more accurate specificity of the minor class (at 25%) was far lower at 0.4


3b) There are a total of 66023 unique CpG coordinates in the dataset. Since Tissue #2 (Islet cells) is the minority class at 25% of data, 25% of that is 16505 CpG coordinates. 1 million reads would mean Tissue #2 has approximately 250,000 reads. 

Coverage_per_target = Total Reads/NUmber of Targets
These 250,000 reads are distributed to across only 16505 PMP's would mean a depth of 15 per CpG. 

Therefore the threshold is 15 reads per biomarker.


3c) The specificity hypothesis
```{r Hypothesis testing 3C}
index <- head(order(test_set_pmp$probability,decreasing = T),n = 10)
top10_cfdna <- test_set_pmp[index,]
index <- head(order(test_set_pmp$probability,decreasing = F),n = 10)
top10_islet <- test_set_pmp[index,]
summary(top10_cfdna[,c(11,13)])
summary(top10_islet[,c(11,13)])
```

As depicted by the boxplot earlier and the relative coverage of Islet cells and cfDNA, higher coverage leads to more certainity and thus higher specificity.






